An algorithm for the optimization of finite element integration loops

### 2016-12-05 TOMS_Paper/assembly.tex
More edits about flops vs performance

Throughout the paper we have reiterated the claim that Algorithm~\ref{algo:gamma} achieves a globally optimal flop count if stronger preconditions on the input variational form are satisfied. We state here these preconditions, in increasing order of complexity.
\begin{enumerate}
\item There is a single monomial and only a specific coefficient (e.g., the coordinates field). This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
\item There is a single monomial, but multiple coefficients are present. Optimality is achieved if and only if all sub-expressions depending on coefficients are structured (see Section~\ref{sec:se-rln}). This avoids ambiguity in factorization, which in turn guarantees that the output of step (7) in Algorithm~\ref{algo:sharing-elimination} is optimal.
\item There are multiple monomials, but either at most one coefficient (e.g., the coordinates field) or multiple coefficients not inducing sharing across different monomials are present. This reduces, respectively, to cases (1) and (2) above.
\item There are multiple monomials, and coefficients are shared across monomials. Optimality is reached if and only if the coefficient-dependent sub-expressions produced by Algorithm~\ref{algo:sharing-elimination} -- that is, the by-product of factorizing test/trial functions from distinct monomials -- preserve structure. 
\end{enumerate}

@@

The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained.

@@

\paragraph{Elasticity}
The results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should. To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}. \texttt{cfO2} finds pre-evaluation profitable in terms of operation count, although it is eventually not applied to avoid exceeding $T_H$. However, running the same experiments with $T_H = \operatorname{size}(\mathrm{L3})$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}. The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the saving in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice. This suggests that our objective function should be improved to handle the cases in which there is a significant gap between potential cache misses and reduction in operation count.

@@

@@

We also tried to relate the achieved performance to the operation counts obtained by each transformation system. We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups are generally massive. For the untransformed, original code, at the source level $37 \cdot 10^9$ operations are counted. However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have aggressively been applied. The actual operation count, however, is definitely larger than the $10^8$ operations performed by \texttt{ufls}. \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}. This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by more than 3$\times$.

>>>>

Throughout the paper we have reiterated the claim that Algorithm~\ref{algo:gamma} achieves a globally optimal flop count if stronger preconditions on the input variational form are satisfied. We state here these preconditions, in increasing order of complexity.
\begin{enumerate}
\item There is a single monomial and only one coefficient (e.g., the coordinates field). This is by far the simplest scenario, which requires no particular transformation at the level of the outer loops, so optimality naturally follows.
\item There is a single monomial, but multiple coefficients are present. Optimality is achieved if and only if all sub-expressions depending on coefficients are structured (see Section~\ref{sec:se-rln}). This avoids ambiguity in factorization, which in turn guarantees that the output of step (7) in Algorithm~\ref{algo:sharing-elimination} is optimal.
\item There are multiple monomials, but either at most one coefficient (e.g., the coordinates field) or multiple coefficients not inducing sharing across different monomials are present. This reduces, respectively, to cases (1) and (2) above.
\item There are multiple monomials, and coefficients are shared across monomials. Optimality is reached if and only if the coefficient-dependent sub-expressions produced by Algorithm~\ref{algo:sharing-elimination} -- that is, the by-product of factorizing test/trial functions from distinct monomials -- preserve structure. 
\end{enumerate}

@@

The rest of the section is organized as follows: we first provide insights into the general outcome of the experimentation; we then comment on the impact of a fundamental low-level optimization, namely autovectorization; finally, we motivate, for each form, the performance results obtained and the relationship with the operation count.

@@

\paragraph{Elasticity}
The results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The main observation is that \texttt{cfO2} never triggers pre-evaluation, although in some occasions it should. To clarify this, consider the test case $\langle \mathrm{nf}=0,\ q=4 \rangle$, in which \texttt{tens/auto} show a considerable speed-up over \texttt{cfO2}. Although pre-evaluation is determined profitable by \texttt{cfO2} in terms of operation count, it is eventually not applied to avoid exceeding $T_H$. However, running the same experiments with $T_H = \operatorname{size}(\mathrm{L3})$ resulted in a dramatic improvement, even higher than that obtained by \texttt{tens}. The reason is that, despite exceeding $T_H$ by roughly 40$\%$, the saving in operation count is so large (5$\times$ in this specific problem) that pre-evaluation would in practice be the winning choice. This suggests that our objective function should be improved to handle the cases in which there is a significant gap between potential cache misses and reduction in operation count.

@@

We now consider the test case $\langle \mathrm{nf}=2,\ q=3 \rangle$ to discuss the relationship between achieved performance and operation count, although analogous considerations apply to the other scenarios. It is difficult to estimate the number of flops executed by the code generated by the FEniCS Form Compiler, since loop-invariant code motion is extensively applied by the underlying compiler. The calculation is instead simple in all other cases: while \texttt{cfO2} performs roughly $2 \cdot 10^5$ operations, \texttt{quad}, \texttt{ufls} and \texttt{cfO1} are much closer to $10^6$ operations. In particular, \texttt{ufls} requires 4.3$\times$ more operations than \texttt{cfO2}, whereas the execution time speed-up is only slightly bigger than 2$\times$. The distance between these two values may be attributed to the memory transfer cost -- especially inserting the $3.6 \cdot 10^3$ non-zero entries of the local element matrix -- which, for reasons already discussed, becomes a bottleneck. \texttt{quad}'s operation count is even lower than that of \texttt{ufls} ($84 \cdot 10^4$ instead of $97 \cdot 10^5$), but as explained earlier, the use of indirection arrays dramatically affects the low level efficiency. Despite requiring roughly $10^5$ fewer operations than \texttt{cfO2}, \texttt{tens} is again affected by the poor low level efficiency of the generated code.

@@

We also tried to relate the achieved performance to the operation counts obtained by each transformation system. We here limit ourselves to consider the case $\langle \mathrm{nf}=0,\ q=4 \rangle$, for which the speed-ups are generally massive. For the original code, at the source level $37 \cdot 10^9$ operations are counted. However, this number is in practice much lower, since by inspection of the assembly code generated by the underlying compiler, it is clear that common sub-expressions elimination and loop-invariant code motion have been applied in aggressive way. The actual operation count, however, is definitely larger than the $10^8$ operations performed by \texttt{ufls}. \texttt{cfO2} brings the operation count down to $3 \cdot 10^7$; that is, more than 3$\times$ lower than \texttt{ufls}. This is completely in line with the drop in execution times, with \texttt{cfO2} outperforming \texttt{ufls} by slightly more than 3$\times$.
